---
title: "Basic Regression"
author: "Math 271"
date: "Spring 2022"
output: 
  html_document: 
    css: lab.css
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(openintro)
library(tidyverse)

```

This material is covered in ModernDive [Chapter 5: Regression](https://moderndive.com/5-regression.html).

## Two Numerical Variables

Suppose we have several quantitative variables measured. The `openintro::satgpa` data set has student information on 1000 college freshmen.

- `SAT[V|M|Sum]` SAT percentile in Verbal|Math|Sum
- `hs_gpa` High School GPA
- `fy_gpa` Freshman year GPA at university



```{r satgpa}
satgpa
```

Plotting is almost always a good idea, and this is especially true when trying to understand the relationship between two variables.

```{r fig1}
ggplot(satgpa) + aes(hs_gpa, fy_gpa) + geom_point() +
  ggtitle("Freshman GPA by HS GPA")
```

The goal of _simple linear regression_ is to find the straight line that best fits the data.

```{r fig1-lm}
ggplot(satgpa) + aes(hs_gpa, fy_gpa) + geom_point() + geom_smooth(method=lm) +
  ggtitle("Freshman GPA by HS GPA")
```

Simple linear regression focuses on exactly two variables:

```{r rows.print=5}
satgpa %>% select(hs_gpa, fy_gpa)
```

One variable must be selected to be a _predictor_ and the other to be the _response_. Here are some other commonly used terms to describe this dichotomy:

- explanatory / independent / cause / $X$
- response / dependent / effect / $Y$

When making a plot, __always__ place the predictor on the horizontal ($X$) axis, and the response on the vertical ($Y$) axis.

In the examples above, the `hs_gpa` is the _predictor_, and `FSGPA` is the _response_. 
When assigning the variables to the $X$ and $Y$ positions, keep in mind what you want to use the model to help you understand.

Consider the two options:

- Use HS grades to understand/predict college grades
- Use college grades to understand/predict HS grades

One of these two options is obviously more useful than the other, given how time works in our universe.

### Notation

Each observation can be thought of as an ordered pair of numbers $(X, Y)$, which we can also think of as locations of points in the Cartesian plane. The graph of all the Cartesian points is the _scatter plot_.

The general equation of a straight line in the plane is \[y = a + bx\]

- $a$ the y-intercept
  + the $y$ value when $x=0$
- $b$ the slope of the line (rise/run)
  + the amount $y$ changes when $x$ increases by 1
  
Given a data point like \((`r (x1y1<-unlist(satgpa[1,5:6]))`)\), we have $X=`r x1y1[1]`$ and $Y=`r x1y1[2]`$. Furthermore, suppose we have a particular equation for a line, say with $a=0.1$ and $b=0.75$,  \[y = f(x) = 0.1 + 0.75 x.\]

```{r}
myline <- function(x){0.1 + 0.75 * x}
myplot <- ggplot(satgpa) + aes(hs_gpa, fy_gpa) + stat_function(fun=myline)
myplot + geom_point(data=slice_head)
```

The line equation will give a _different_ number than the y-value of the data point. The height of the line is the _prediction_ of the y-value when $X=`r x1y1[1]`$.

\[ f(`r x1y1[1]`)= 0.1 + 0.75 \cdot `r x1y1[1]` = `r myline(x1y1[1])`\]

- $X=`r x1y1[1]`$ _observed_ explanatory value
- $Y = `r x1y1[2]`$ _observed_ response value
- $\hat Y=`r myline(x1y1[1])`$ _predicted_ response value


The difference between the observation and the prediction is called the __residual__ or the __error__ for that data point. 

\[ E = Y - \hat Y= `r x1y1[2] - myline(x1y1[1])`\]

- $E = Y - \hat Y$ the _error_ in the prediction

The residual is the vertical distance from the line to the point
  
```{r}
myplot + geom_point(data=slice_head) + geom_segment(aes(xend=hs_gpa, yend=myline(hs_gpa)), data=slice_head)
```


Each data point has a different value for $X, Y, \hat Y, E$, the residuals are positive when $Y>\hat Y$, and the point lies above the line on the graph. The residual is negative when the point is below the line, and $Y<\hat Y$.

```{r, cache=TRUE}
ggplot(satgpa %>% slice_sample(n=10)) + aes(hs_gpa, fy_gpa) + stat_function(fun=myline) + 
  geom_point() + geom_segment(aes(xend=hs_gpa, yend=myline(hs_gpa)))

```

If we want to refer to the values associated with row $i$ in the data set, we add a subscript to the letters.
The values we calculated above are associated with the first row of data, so they would be $X_1, Y_1, \hat Y_1, E_1$.

If our data is observational, we do not have control over any of these numbers. However, in a designed experiment, we are often able to directly control the $X$ values, e.g., _How much fertilizer will I apply to this plant?_

## Estimating the linear regression model

With real data, we do not know the equation of the real line; $a$ and $b$ are unknown. Our goal is to _estimate_ them from the data

As usual, we add hats to indicate estimated values, $\hat a$, $\hat b$, will refer to the values for the best-fit line for a data set.

Obtaining the usual estimates of these quantities is done with the `lm` command:

```{r}
## lm(Y ~ X, data)
lm(fy_gpa ~ hs_gpa, data=satgpa)
```

The first argument to `lm` is a _model `formula`_. For simple linear models the formula is `response ~ predictor`, but we can fit many different models using this command by changing the formula. 

The output of the `lm` command is very useful to capture in a variable for further use:

```{r}
sat.lm <- lm(fy_gpa ~ hs_gpa, satgpa)
```

The _Coefficients_  give the estimated values $\hat a$ and $\hat b$. These are accessible as an object with the `coef` command
```{r}
coef(sat.lm)
```

Having these values available is useful, for example:
```{r}
ab <- coef(sat.lm) ## icept is ab[1] and slope is ab[2]
ggplot(satgpa) + aes(hs_gpa, fy_gpa) + geom_point(alpha=0.2) +
  geom_abline(intercept=ab[1] , slope=ab[2] ) + 
  ggtitle("Freshman GPA by HS GPA")
```

However, for pure plotting purposes, the easiest method is to use `geom_smooth(method="lm")`
```{r}
ggplot(satgpa) + aes(hs_gpa, fy_gpa) + geom_jitter(alpha=0.2, width=0.01) + geom_smooth(method="lm") + 
  ggtitle("Freshman GPA by HS GPA")
```


## Commands for fitted `lm` objects

- `summary` gives a lot more information about the linear regression problem. Most of the output here will be explained in forthcoming lectures.
```{r}
summary(sat.lm)
```

- `coef` gives the vector of regression coefficients
- `predict` gives the vector of predicted $\hat Y$ values (among other things)
- `residuals` gives the vector of residuals

```{r}
satgpa %>% select(hs_gpa, fy_gpa) %>% mutate(yhat=predict(sat.lm), e = residuals(sat.lm))
```

We can verify the relationship between the observations, the residuals and the predictions

```{r}
all.equal(satgpa$fy_gpa, (predict(sat.lm) + residuals(sat.lm)), check.names=FALSE)
```


If you want to use the model to predict on some new $X$ values, we can make a new data frame and give it as a second option to `predict`

```{r}
new.xvals <- data.frame(hs_gpa=c(3.3, 1.5, 4.0))
new.yhat <- predict(sat.lm, new.xvals)
new.xvals %>% mutate(new.yhat)
```

## Help and Object-Oriented programming

If we look at the help for these commands, e.g. `?predict`, it's not very detailed. For example, it doesn't tell you anything about the ability to predict on new x values.

The basic `predict` command is what R terms a _generic function_. It can do many different things based on what _class_ of object is given to it.

Every object in R has a _class_ that you can check using the `class` command

```{r}
class(1)
class("a")
class(sat.lm)
```

The `predict` command has different behaviors defined for many different classes input objects. We can see what special behaviors are defined for a command by asking about its `methods`.

```{r}
methods(predict)
```

We see above that there is a method `predict.lm` which will be used when we give predict an input of class `"lm"`. If we want to know the details of this special method, we can look up the help on that function: `?predict.lm`. This reveals the second `newdata` argument available to us in this situation.

### Special CS Topic

The Object Oriented system in R is implemented with _function polymorphism_, which is a different from how OO is implemented in languages like Python and C++. Class methods are associated with a _generic function_ instead of being associated with the class definition. Many basic commands in R are registered as _generic_. Generic function definitions are very short, usually consisting of only a call to `UseMethod`.
```{r}
summary
```

If you want to implement a special method for a generic command that applies to class `"foo"`, you simply define a new function named, e.g. `summary.foo`. You can change the _class_ of an object by assigning `class(object)` a new value.

```{r}
x <- "an object"
summary(x)

summary.foo <- function(x){"this is the summary.foo method"}
class(x) <- "foo"
summary(x)
```
## Excercises

Create a solution file `PS06_solution.Rmd` and address the following questions with code, plots, or text answers, as appropriate. When complete, commit your solution `.Rmd` and `.html` files to your repository and push them to Github. Request a code review from @grady on the Feedback Pull Request.

1. Load the `openintro` package and become acquainted with the `evals` data set. Read the description in the help files, and use your skills to answer these questions:
    - How many rows and columns does the data have, what is the observational unit described in each row? Are there any missing data values present?
    - Across all classes, find the total number of students enrolled, the number of completed evaluations, and the percent of evaluations that were completed. Also, compute the mean of the `cls_perc_eval` column. Why are these not the same?
    - What is the mean number of courses in the data per professor.
    
    
2. Make a scatter plot with the `bty_avg` variable on the horizontal axis, and `score` on the vertical axis. Use `geom_jitter` to help alleviate the over-plotting problem. Play around with the `alpha` and `width`/`height` parameters until you get a nice informative picture. (I recommend starting the width and height at 0.01 and increasing slowly until it looks good.) Be sure to set the axis labels to something more informative. 

3. Using your plot from the previous questions as the $X$ and $Y$ variables in linear regression, use `lm` to fit the regression model to the `evals` data set. Store the result in the `bty_score` object. What are the intercept and slope coefficients for the best fitting model?
    + A basic solution will simply print code output that has the slope and intercept in it. 
    + A better solution will report the slope and intercept in a textual description of the results.
    + A great solution will use in-line R code chunks to directly inject the coefficient values into the text. (Print a reasonable number of sig figs.)
    + Wizard level: Create a LaTeX equation describing the fitted line, with the coefficient values injected by inline R code chunks.

4. Create a new data frame named `evals_resid`, with the columns `bty_avg`, `age`, `score` from the original data, and new columns `score_pred` and `score_age` with predictions and residuals from the linear model.

5. Create a histogram of the residuals. Do the residuals appear to be normally distributed, or skewed?

6. Create a _residual_ plot by plotting the predicted values on the horizontal axis and the residual values on the vertical axis. 
    
7. Fit a new linear regression model using `bty_avg` to predict `age`. Store the result in `bty_age`. Use this new model to add columns `age_pred` and `age_resid` to the `evals_resid` data set.

8. Plot `age_resid` on the horizontal axis, and `score_resid` on the vertical axis. 

At this point, the final plot is just an exercise in coding, but this plot is known as a _partial regression_ or _added variable_ plot for `age`. It gives some indication about what additional information about the `score` might be contained in the `age` variable, beyond what `bty_avg` already tells us. In this case, the lack of any discernible pattern in the plot indicates that there is not much new information.